---
title: 'Ecology Lab: Prairie Dog Analysis'
date: "Fall 2018"
output: html_document
---
*******

Script for EBIO 2040 Data Analysis II- Analyzing the Prairie Dog Data
Prairie dog effects on vegetation on a large scale (on and off a prairie dog colony)
Updated by: Kim Vincent 23 Jun 2018
Edited by: Kelsey Elwood 30 Sept 2018

***********
# Preparation

This script is designed to introduce students to using R for statistical analysis and covers the following: 
1. Setting a working directory
2. Uploading data files
3. Exploring dataframes for structure and content
4. Visualizing data with plots
5. Adjusting plots 
6. Applying a t-test

## Setting your working directory

**Step 1:** Create or locate a folder to use as working directory. Remember, a _working directory_ is the folder that contains your data and your R scripts. Save the folder to your desktop or another location that you'd prefer. Do NOT use the "Downloads" folder.  
- If you are using the classroom computers, I recommend using a thumbdrive as your working directory so that you can use it on other computers.  
- Do not put spaces in the title of the folder name or any folders located "above" your working directory. 

**Step 2:** Determine the current working directory. 
Typing the folowing line of code,  which translates to "get working directory", 
displays the current working directory of the computer in use (including the user pathname).  
```{r}
getwd() 

```

**Step 3:** Change your working directory to the folder you decided on in Step 1. 
Typing the folowing line of code, which translates to "set working directory", assigns the specified folder as the working directory. 
To find the location, go to this file in the finder, click on file tab, choose "get info" and copy (command + C) the location noted under "Where:". Paste (command + v) the location inside quotations in the setwd() command:
```{r}
setwd("ENTER_FOLDER_LOCATION") # e.g. "/Users/elwoodk/Desktop/ecology-lab5-data"

```

**Step 4:** Check if the working directory has changed by typing in the following again: 
```{r}
getwd()

```

## Load all required packages

Because we want to add error bars to graphs, we need to load the package "plotrix"
```{r}
library(plotrix)
# If you get an error, you might need to install the package and THEN load it using `library()`. To do that, run the following line of code (without the hashtag in front):
# install.packages("plotrix")
# library(plotrix)
```

# Large Scale Experiment: Comparing two categorical variables. 

## Import .csv files of data

The prairie dog experiment has two datasets: 1) Large Scale and 2) Small Scale
First, let's work with the large scale dataset.

**Step 1:** Save a copy of the Large Scale experiment dataset as a .csv file in your working directory. 
(You will need a .csv file and not an .xlsx because R and Excel are not friends.)

**Step 2:** Create "objects" of the spreadsheets so that we can easily reference the spreadsheets for analysis throughout the R script. Objects are created when we use the notation "<-". The new object name is on the left of the arrow and represents the output of the function on the right of the arrow (in this case, `read.csv()`)

First, let's look at the semesters separately, by loading the files for spring 2018 and fall 2017 only. We can load the spreadsheets (in csv format) by using just the name of the file (as seen below) because the files are in our working directory, so R already knows to look in that folder for the files. If we did not set the working directory, we would need to include all the names of the folders that housed our file.

```{r}
LargeS18.dat <- read.csv("Large_Scale_S18.csv") 
LargeF17.dat <- read.csv("Large_Scale_F17.csv")  
```

Now, upload the "Large_Scale_All.csv" file, which includes data from all semesters, and name the object "Large.dat". *Note that you will need to type in the name of the file.* 
```{r}
LargeAll.dat <- read.csv("ENTER_FILE_NAME") 

```


## Exploring your data

A couple example functions to use to visualize various parts of the dataset. 

The function "View()" opens a tab that shows you the full spreadsheet.
```{r}
View(LargeAll.dat)

```

The function "str()" shows the structure of the object. *Make sure that categorical variables are called "factor" and that continuous variables are called "int" (i.e. integer) or "num" (i.e. numeric).* 
```{r}
str(LargeAll.dat)

```

The function `summary()` provides information on the mean, median, and distributions of the data:
```{r}
summary(LargeAll.dat)
summary(LargeF17.dat)
summary(LargeS18.dat)

```

## Plotting Data with Categorical Independent Variables

Always visualize your data as a first step before you run statistical tests. 

We will cover two kinds of plots: the boxplot and the bar graph

### THE BASIC BOX PLOT

- Boxplots display the treatment *medians*, while bargraphs display the *mean.*   
- Boxplots are useful to visualize your data because they display outliers (small open circles) and if the notches overlap, there is a good chance the means are significantly different.  
- Because t-tests are based on comparing means, a bar graph should be used in conjunction with a discussion of means within a manuscript. 

Let's start with the boxplot to visualize our data. 
Use the following line of command to plot the abundance of vegetation tallied for the on-colony site and the off-colony site. 

```{r}
boxplot(Total_Veg~PD_Colony, # This line translates to: "Total_Veg column is dependent on PD_Colony column"
        data = LargeAll.dat, #The columns come from the spreadsheed we called "LargeAll.dat"
        notch = TRUE, # Include the notch, which represents the 95% confidence around the median
        main = "Vegetation Abundance\nOn and Off a Prairie Dog Colony", # The main title. You can use the notation "\n" to separate the title into multiple lines.
        ylab = "Vegetation Abundance", # The y-axis label
        xlab = "Prairie Dog Colony", # The x-axis label
        col = c("olivedrab1", "snow3"), # The colors of the two boxes. The c() represents a list of colors. For more color options, see http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
        cex.axis = 1, # The size of the numbers on the axis
        cex.main = 1.5, # The size of the main title
        cex.lab = 1.3) # The size of the text in the axis labels
```

### BAR GRAPH 

**Step 1:** Calculate the mean values for each group and assign the output a meaningful name. The mean will be indicated as the top of the bar in the bar chart.
```{r}
veg_abundance_mean <- aggregate(Total_Veg ~ PD_Colony, # Do math on the "Total_Veg" column for each group in the "PD_Colony" column
                                data = LargeAll.dat, # Using the dataset we named "LargeAll.dat"
                                FUN = mean) # The math (i.e. function) that we will apply is the "mean"

# view the new object called "veg_abundance_mean" by running the object name:
veg_abundance_mean
```

**Step 2:** Calculate the standard error (SE) for each group, which will be shown in the "error bars" or little T's that go above and below the mean on the bar chart. SE = sqrt(variance/n) = standard deviation/sqrt(n), where n is sample size.

```{r}
# Create a function that calculates SE:
calcSE <- function(x){sd(x)/sqrt(length(x))}

# Use `aggregate()` function to calculate SE for each group in our dataset:
veg_abundance_SE <- aggregate(Total_Veg~PD_Colony, # Do math on the "Total_Veg" column for each group in the "PD_Colony" column
                              data = LargeAll.dat, # Using the dataset we named "LargeAll.dat"
                              FUN = calcSE) # The math (i.e. function) that we will apply is the "calcSE" that we created above

# view new object called "veg_abundance_SE" by running the object name:
veg_abundance_SE
```

**Step 3:** Create a bar plot with both mean and SE represented for each group. *Note how the barplot command is included inside the plotCI command.*
```{r}
plotCI(barplot(height = veg_abundance_mean[,2], # The height is the mean, which is found in the second column [,2] of the object we named "veg_abundance_mean"
               beside = TRUE, # Plot both bars side-by-side
               ylim = c(0, 50), # Lower and upper limits for the y axis -- You may need to adjust this!
               ylab = "Vegetation Abundance", # y-axis label
               xlab = "Transect Type", # x-axis label
               col = c("limegreen","lemonchiffon2"), # colors of the 2 bars
               names.arg = veg_abundance_mean[,1], # The names of the two columns are located in the first column [,1] of the object we named "veg_abundance_mean" 
               cex.axis = 1, # Size of the numbers on the axes
               cex.names = 1), # Size of the names on the x-axis
       veg_abundance_mean[,2], # Place the confidence intervals (SE) around the mean, which is located in the second column [,2] of the object we named "veg_abundance_mean"
       uiw = veg_abundance_SE[,2], # The width of the confidence interval (SE) is located in the second column [,2] of the object we named "veg_abundance_SE"
       add = TRUE, # add the confidence intervals to the barplot
       pch = NA) # remove a circle that would otherwise print on the plot (we don't need the circle, so we can put "NA")
```

The appropriate scaling for the limit of the y axis is just larger than the mean + SE. Add together the mean + SE and round up to the next round number. *Change the code above to improve the scaling on the y-axis.*

## DATA ANALYSIS: T-Test

Now for the real test! To determine if we are confident enough to reject the null hypothesis that the mean vegetation abundance is the same on-colony and off-colony, we need to run a t-test. The goal of a t-test is to compare the means of a dependent variable between two categories of an independent variable. 

Since the independent variable for the Large Scale experiment is composed of two categorical variables, we use a t-test for the analysis. To run a t-test in R, use the *function* t.test() followed by the response (dependent) variable as affected by the predictor (independent) variable. 

Run a t-test to determine if the amount of *Total_Veg* abundance differs on and off colonies. 
Note the:
*p-value*: If the values is less than 0.05, than that means that there is a "significant difference" between the two groups.
*t-value*: We will report this when we report our statistics. The t-value is what the computer used to calculate the p-value. For now, don't worry about how it is calculated.
*df*: We will report this when we report our statistics. "df" stands for degrees of freedom. For now, don't worry about what this means or how it is calculated.
```{r}
t.test(DEPENDENT ~ INDEPENDENT, 
       data = DATA_OBJECT)

```

In addition to the output from the t-test, we will also want to report the means and SE of the two groups. We can do that by "printing" the object names that we created earlier
```{r}
veg_abundance_mean
veg_abundance_SE
```

*********
Let's explore another variable. How would you change the code above to look at whether the amount of *"Grass"* abundance differs on and off colonies?

```{r}
# Step 1: Visualize with a boxplot
boxplot(DEPENDENT ~ PD_Colony, # Use Grass as the DEPENDENT variable. Note that Grass is capitalized, which matches the way it appears in the LargeAll.dat file
        data = LargeAll.dat, #The columns come from the spreadsheed we called "LargeAll.dat"
        notch = TRUE, # Include the notch, which represents the 95% confidence around the median
        main = "Vegetation Abundance\nOn and Off a Prairie Dog Colony", # The main title. You can use the notation "\n" to separate the title into multiple lines.
        ylab = "Grass Abundance", # The y-axis label
        xlab = "Prairie Dog Colony", # The x-axis label
        col = c("olivedrab1", "snow3"), # The colors of the two boxes. The c() represents a list of colors. For more color options, see http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
        cex.axis = 1, # The size of the numbers on the axis
        cex.main = 1.5, # The size of the main title
        cex.lab = 1.3) # The size of the text in the axis labels

# Step 2: Use `aggregate()` function to calculate the mean for each group in our dataset:
CHOOSE_OBJECT_NAME <- aggregate(DEPENDENT ~ PD_Colony, # Do math on the DEPENDENT column for each group in the "PD_Colony" column
                                data = LargeAll.dat, # Using the dataset we named "LargeAll.dat"
                                FUN = mean) # The math (i.e. function) that we will apply is the "mean"

# Step 3: Use `aggregate()` function to calculate the standard error (SE) for each group in our dataset and store the output as an object
CHOOSE_OBJECT_NAME <- aggregate(DEPENDENT ~ PD_Colony, # Do math on the DEPENDENT column (make sure to get the spelling and capitalization exactly the same as it appears in the csv file!) for each group in the "PD_Colony" column
                              data = LargeAll.dat, # Using the dataset we named "LargeAll.dat"
                              FUN = calcSE) # The math (i.e. function) that we will apply is the "calcSE" that we created above

# Step 4: Barplot
plotCI(barplot(height = OBJECT_WITH_MEAN_VALUES[,2], # The height is the mean, which is found in the second column [,2] of the object we named above that holds the mean values (step 2)
               beside = TRUE, # Plot both bars side-by-side
               ylim = c(0, 50), # Lower and upper limits for the y axis -- You may need to adjust this!
               ylab = "ENTER LABEL FOR Y-AXIS HERE", # y-axis label
               xlab = "Transect Type", # x-axis label
               col = c("limegreen","lemonchiffon2"), # colors of the 2 bars
               names.arg = OBJECT_WITH_MEAN_VALUES[,1], # The names of the two columns are located in the first column [,1] of the object that holds the mean values
               cex.axis = 1, # Size of the numbers on the axes
               cex.names = 1), # Size of the names on the x-axis
       OBJECT_WITH_MEAN_VALUES[,2], # Place the confidence intervals (SE) around the mean, which is located in the second column [,2] of the object that holds the mean values
       uiw = OBJECT_WITH_SE_VALUES[,2], # The width of the confidence interval (SE) is located in the second column [,2] of the object that holds the SE values (step 3)
       add = TRUE, # add the confidence intervals to the barplot
       pch = NA) # remove a circle that would otherwise print on the plot (we don't need the circle, so we can put "NA")

# Step 5: Compare the means of the on- versus off-colony grass abundances
t.test(DEPENDENT ~ PD_Colony, 
       data = LargeAll.dat)

# Step 6: Review the means and SE values: Replace the text below with the object names you chose in steps 2 and 3
OBJECT_WITH_MEAN_VALUES
OBJECT_WITH_SE_VALUES
```

*********************************

# Small Scale Experiment: Comparing two continuous variables. 

## Import .csv files of data:
```{r}
# Load the "Small_Scale_All.csv" file:
SmallAll.dat <- read.csv("TYPE_FILE_NAME_HERE.csv") 

```

## Exploring the data you uploaded (the spreadsheet we loaded is known as a "dataframe" in R)

A couple example functions to use to look at various parts of the dataset. These are useful to visualize your data. 

```{r}
# View the entire dataset in a new tab
View(SmallAll.dat)

# Display the first several rows of the object (the dataframe, SmallAll.dat).
head(SmallAll.dat)

```

Display the structure of the object, SmallAll.dat.
```{r}
str(SmallAll.dat)

```

## Exploring the data using a graph (visualizing your data)
This time we have continuous data for both the independent and the dependent variables. We can use the simple `plot()` function to look at our data. 
```{r}
plot(Total_Percent_Veg ~ Dist_to_Burrow, # Plot the value in the "Total_Percent_Veg" column as a function of the value in the "Dist_to_Burrow" column (i.e. Grass = y-axis, Dist_to_Burrow = x-axis)
     data = SmallAll.dat, # The name of the object that holds our spreadsheet
     bty = "l", # Include lines on the x- and y-axes to look like an "L" (other options include "c", "7", and "o")
     ylab = "Percent Vegetation Cover", # y-axis label
     xlab = "Distance to Nearest Burrow (m)", # x-axis label
     cex.lab = 1.5, # size of the axes labels (text)
     cex.axis = 0.75) # size of the axis numbers
```

*What do you think the relationship is between the distance to burrow and the percent of vegetation cover? Is there more vegetation close to or far from a burrow?*

The relationship is difficult to see, isn't it? To assess whether or not there is a significant relationship between the variables, we have to run a regression test using the `lm()` function. We will store the output of the regression test as an object called "TotalVeg_All.lm"
```{r}
TotalVeg_All.lm <- lm(Total_Percent_Veg ~ Dist_to_Burrow, # Run a regression (aka "linear model") where Total_Percent_Veg depends on the Dist_to_Burrow
                     data = SmallAll.dat) # Use the SmallAll.dat dataset
```

When running a regression, we want to note a couple variables: 
- The *p-value* (Pr(>|t|) of the INDEPENDENT variable (in this case "Dist_to_Burrow"), which will tell us if there is a "significant" relationship between the variables. The relationship is ONLY significant if the p-value is BELOW 0.05. 
- The *multiple R-squared*, which represents how much variation there is in the data around the regression line. Values close to 1.0 are great and represent that all the data lines up perfectly on our regression line. This would mean that the independent variable perfectly explains all the values found for the dependent variable. R-squared values close to 0 are bad and mean that many data points are located far from our regression line. This would mean that the independent variable only explains some of the variation in the dependent variable (and that there might be other, unmeasured factors that could influence the value of the dependent variable). In ecology, we often aim for R-squared values around 0.2-0.3 (but bigger is always better in this case!) 

To see the results of the regression, we can use the `summary()` function
```{r}
summary(TotalVeg_All.lm) 
```

*What is the p-value (Pr(>|t|) of the regression test? Is there a "significant" relationship between the distance to the nearest burrow and the total vegetation? Is there a lot of variation in the data?*

If there is a significant relationship (p < 0.05), then we will want to represent that by adding the regression line to our plot. If the relationship is non-significant (p > 0.05), you will want to omit the line because there really isn't a relationship. 

Let's add a line of best fit to visualize the relationship.
```{r}
abline(TotalVeg_All.lm)

```

### Adjusting the code to look at other vegetation categories in the SmallAll.dat dataset

Plot the data for *bareground*
```{r}
# Initial plot for bareground:
plot(Bareground ~ Dist_to_Burrow, 
     data = SmallAll.dat, 
     bty = "l", 
     ylab = "Percent Bareground", 
     xlab = "Distance to Nearest Burrow (m)",
     cex.lab = 1, cex.axis = 0.75)

# Regression test: Is the relationship significant for bareground as a function of distance to burrow?
Bareground_All.lm <- lm(Bareground ~ Dist_to_Burrow, data = SmallAll.dat)

# Review regression output
summary(Bareground_All.lm)

# Add regression line (only if significant)
abline(Bareground_All.lm) 
```


*How would you change the code to look at Grass?*
```{r}
# Initial plot for grass:



# Regression test for grass:



# Summary of regression test for grass:




# Add `abline()` ONLY if there is a significant relationship:




```


### Graphing and testing results for other years

We have spreadsheets (csv files) for each semester in our working directory. Update the following code to look at the data for each semester individually.

*Only Fall 2017*
```{r}
# Load data
SmallF17.dat <- read.csv("ENTER_FILENAME_HERE.csv")

# Create plot
plot(Percent_Veg ~ Dist_to_Burrow, 
     data = SmallF17.dat, 
     bty = "l", 
     ylab = "Percent Vegetation Cover: Fall 2017", 
     xlab = "Distance to Nearest Burrow (m)",
     cex.lab = 1, cex.axis = 0.75)

# Run regression
Percent_F17.lm <- lm(Total_Percent_Veg ~ Dist_to_Burrow, 
                     data = SmallF17.dat)

# Review regression output
summary(Percent_F17.lm)

# Add regression line (only if significant)
abline(Percent_F17.lm)

```


*Only Spring 2018*
```{r}
# Load data
SmallS18.dat <- read.csv("ENTER_FILENAME_HERE.csv")

# Create plot
plot(Total_Percent_Veg ~ Dist_to_Burrow, 
     data = DATA_OBJECT_NAME, 
     bty = "l", 
     ylab = "LABEL FOR Y-AXIS (i.e. DEPENDENT VARIABLE)", 
     xlab = "Distance to Nearest Burrow (m)",
     cex.lab = 1, cex.axis = 0.75)

# Run regression
Percent_S17.lm <- lm(Total_Percent_Veg ~ Dist_to_Burrow, 
                     data = DATA_OBJECT_NAME)

# Evaluate regression results
summary(REGRESSION_OBJECT)

# Add regression line if significant
abline(REGRESSION_OBJECT)
```


*Only Fall 2018*
```{r}
# Load data


# Create plot


# Run regression


# Evaluate regression results


# Add regression line if significant


```

